CuO Analysis Details

### 2019-06-07 paper/draft.tex
methods DMD and FNDMC done

\section{Methods}
The DMD procedure allows us to develop low-energy effective theories for physical systems in a systematic manner beginning with the \textit{ab-initio} Hamiltonian:
\begin{equation}
H_\text{ab} = -\frac{1}{2} \sum_{i} \nabla_i^2 - \sum_{i,I}\frac{Z}{|r_i - R_I|} + \sum_{i<j}\frac{1}{|r_i - r_j|}
\label{eq:Hab}
\end{equation}
where $i$ indicates electrons and $I$ nuclei, and the unit of energy is Hartree (Ha).
In this paper we will be working under the Born-Oppenheimer approximation and core electrons will be treated using pseudo-potentials.
The procedure defines a method for downfolding the \textit{ab-initio} energy functional and Hilbert space onto an effective energy functional defined on a low-energy subspace of the Hilbert space 
\begin{equation}
(E_\text{ab}, \mathcal{H}) \xrightarrow{\text{DMD}} (E_\text{eff}, \mathcal{LE} \subset \mathcal{H}), \text{ s.t. }
\forall |\Psi\rangle \in \mathcal{LE}, \ E_\text{eff}[\Psi] - E_\text{ab}[\Psi] = \epsilon[\Psi] + E_0.
\label{eq:DMD}
\end{equation} 
The energy functionals can be written as expectations over Hamiltonian operators $E_\text{ab} = \langle \Psi | H_\text{ab} |\Psi \rangle$, $E_\text{eff} = \langle \Psi | H_\text{eff} |\Psi \rangle$, $\epsilon$ represents the error in our effective theory, and $E_0$ is a constant energy shift.
Broadly the DMD method consists of four steps which will be discussed in detail later: 
A) Defining the space of low-energy excitations which $E_\text{eff}$ should be able to describe accurately.
B) Sampling, at minimum, a set of states in $\mathcal{H}$ which probe the chosen low-energy excitations in a statistically independent manner. The selected states need not be eigenstates. 
C) Constructing a set of candidate models which are linear in reduced density matrix (RDM) descriptors with variable coefficients $\{c\}$
\begin{equation}
E_\text{eff} = \sum_k c_k d_k[\Psi]\ + E_0,\ d_k[\Psi] = \langle \Psi | \hat{d}_k |\Psi \rangle.
\label{eq:Eeff}
\end{equation}
Here $\hat{d}_k$ are Hermitian operators typically composed of 1- or 2-RDM elements. 
Examples of possible $\hat{d}_k$ are number operators $\hat{n}$ and exchange operators $\vec{S}_i \cdot \vec{S}_j$.
Additionally, a single particle basis must be built to express RDM elements on and should be chosen such that variations among states in $\mathcal{LE}$ are mostly described by variations in the projection of these states onto the basis.
D) Using the samples to fit the coefficients $\{c\}$ by linear regression such that $E_\text{eff}$ satisfies the conditions in \eqref{eq:DMD} with the smallest error $\epsilon$ possible. 
The fitting procedure is typically an ordinary linear regression, however, one may use a custom cost function in the regression as well. 
In order to conduct the linear regression for a functional like \eqref{eq:Eeff}, one needs to calculate for each state sampled from $\mathcal{LE}$ the expectation values of $H_\text{ab}$ and every descriptor $\{\hat{d}_k\}$ used in the regression.
In principle, if conducted with no approximations, the DMD method is rigorously guaranteed to obtain exact results, namely $\epsilon[\Psi] = 0$.

\pagebreak
peepoo

@@

Based on anion photoelectron spectroscopy (APES) experiments\textbf{[ref]} on the CuO molecule we define our low-energy space $\mathcal{LE}$ as the space of states with 9 to 10 electrons in the Cu 3d orbitals. 
Mathematically we can express this space as the span of some eigenstates within our Hilbert space
\begin{equation}
\mathcal{LE} = \text{Span(}\{ |\Psi \rangle | \langle \Psi | n_{3d} | \Psi \rangle \ge 9,\ H|\Psi\rangle = E |\Psi\rangle \}\text{)}.
\end{equation}
The APES measurements indicate that the lowest energy excitations in the CuO molecule are between the O p-orbitals and Cu 4s orbitals, leaving around 10 electrons in the Cu 3d orbitals. 
Bonding between the O p and Cu 3d orbitals lead to significant variations in the Cu 3d occupation, indicating the importance of including excitations out of the Cu 3d orbitals. 
This is corroborated by APES measurements indicating that the lowest energy eigenstates with 9 electrons in the Cu 3d orbitals are only $\sim$ 0.2 eV higher in energy than the second excited state of the system with 10 electrons in the Cu 3d orbitals. 
Our particular focus will be on describing most accurately the properties and energies of eigenstates up to 2eV, as higher energy states form a nearly continuous block of states which include vibrational modes.

>>>>

\section{Methods}
\subsection{Density Matrix Downfolding (DMD)}
The DMD procedure allows us to develop low-energy effective theories for physical systems in a systematic manner beginning with the \textit{ab-initio} Hamiltonian:
\begin{equation}
H_\text{ab} = -\frac{1}{2} \sum_{i} \nabla_i^2 - \sum_{i,I}\frac{Z}{|r_i - R_I|} + \sum_{i<j}\frac{1}{|r_i - r_j|}
\label{eq:Hab}
\end{equation}
where $i$ indicates electrons and $I$ nuclei, and the unit of energy is Hartree (Ha).
In this paper we will be working under the Born-Oppenheimer approximation and core electrons will be treated using pseudo-potentials.
The procedure defines a method for downfolding the \textit{ab-initio} energy functional and Hilbert space onto an effective energy functional defined on a low-energy subspace of the Hilbert space 
\begin{equation}
(E_\text{ab}, \mathcal{H}) \xrightarrow{\text{DMD}} (E_\text{eff}, \mathcal{LE} \subset \mathcal{H}), \text{ s.t. }
\forall |\Psi\rangle \in \mathcal{LE}, \ E_\text{eff}[\Psi] - E_\text{ab}[\Psi] = \epsilon[\Psi] + E_0.
\label{eq:DMD}
\end{equation} 
The energy functionals can be written as expectations over Hamiltonian operators $E_\text{ab} = \langle \Psi | H_\text{ab} |\Psi \rangle$, $E_\text{eff} = \langle \Psi | H_\text{eff} |\Psi \rangle$, $\epsilon$ represents the error in our effective theory, and $E_0$ is a constant energy shift.
Broadly the DMD method consists of four steps which will be discussed in detail later: 
A) Defining the space of low-energy excitations which $E_\text{eff}$ should be able to describe accurately.
B) Sampling, at minimum, a set of states in $\mathcal{H}$ which probe the chosen low-energy excitations in a statistically independent manner. The selected states need not be eigenstates. 
C) Constructing a set of candidate models which are linear in reduced density matrix (RDM) descriptors with variable coefficients $\{c\}$
\begin{equation}
E_\text{eff} = \sum_k c_k d_k[\Psi]\ + E_0,\ d_k[\Psi] = \langle \Psi | \hat{d}_k |\Psi \rangle.
\label{eq:Eeff}
\end{equation}
Here $\hat{d}_k$ are Hermitian operators typically composed of 1- or 2-RDM elements. 
Examples of possible $\hat{d}_k$ are number operators $\hat{n}$ and exchange operators $\vec{S}_i \cdot \vec{S}_j$.
Additionally, a single particle basis must be built to express RDM elements on and should be chosen such that variations among states in $\mathcal{LE}$ are mostly described by variations in the projection of these states onto the basis.
D) Using the samples to fit the coefficients $\{c\}$ by linear regression such that $E_\text{eff}$ satisfies the conditions in \eqref{eq:DMD} with the smallest error $\epsilon$ possible. 
The fitting procedure is typically an ordinary linear regression, however, one may use a custom cost function in the regression as well. 
In order to conduct the linear regression for a functional like \eqref{eq:Eeff}, one needs to calculate for each state sampled from $\mathcal{LE}$ the expectation values of $H_\text{ab}$ and every descriptor $\{\hat{d}_k\}$ used in the regression.
In principle, if conducted with no approximations, the DMD method is rigorously guaranteed to obtain exact results, namely $\epsilon[\Psi] = 0$.

\subsection{Fixed-node diffusion Monte Carlo (FN-DMC)}
When downfolding $H_\text{ab}$ we will use FN-DMC to access states in $\mathcal{LE}$ and accurately calculate their \textit{ab-initio} energies and RDM elements.
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a Hamiltonian given some initial trial wave function.
Consider a trial wave function $|\psi_T\rangle$ and the Hamiltonian $H_\text{ab}$ with ground state $|\phi_0\rangle$. We apply the projector $e^{-\tau H_\text{ab}}$ as $\tau \rightarrow \infty$ to $|\psi_T \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau H_\text{ab}} |\psi_T\rangle 
\equiv \lim_{\tau \rightarrow \infty} |\psi_\text{DMC}(\tau)\rangle \propto \langle \phi_0|\psi_T\rangle |\phi_0\rangle,
\end{equation}
projecting out the ground state as long as the trial wave function we choose is not orthogonal to the ground state. 
The stochastic implementation involves moving samples from the trial function $\psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H_\text{ab} - E_T)} | R^\prime \rangle$. Since $H_\text{ab} = T + V$, kinetic and potential energy terms, the Green function is approximated by a Trotter expansion $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle \sim \Big[e^{-d\tau(V(R) + V(R^\prime))/2} \langle R| e^{-d\tau(T - E_T)}|R^\prime \rangle + O(d\tau^2) \Big]^N $ where $d\tau = \tau/N$.
This expansion can be interpreted as an interative procedure where samples are moved N times with a small timestep $d\tau$ until convergence.
The constant $E_T$ is a trial energy used to control the normalization of $\psi_\text{DMC}(\tau, R)$ and is updated at each move.
This implementation, however, suffers from a fermion sign problem. 
We deal with this sign problem through a fixed-node approximation, where the nodal surface of $\psi_\text{DMC}(\tau, R)$ is forced to match that of the initial trial wave function for all $\tau$.
This approximation makes FN-DMC variational, and will only return the exact ground state of $H_\text{ab}$ if the nodal surfaces of $|\psi_T\rangle$ and $|\phi_0\rangle$ are identical.
We make use of this variational principle to access low-energy states within $\mathcal{H}$ by choosing trial wave functions with nodes that differ from the nodes of $|\phi_0 \rangle$.
To calculate the expectation value of an operator $\hat{O}$ in FN-DMC we will use a mixed estimator $\langle \psi_{DMC}(\tau) |\hat{O} | \psi_T \rangle/\langle \psi_\text{DMC}(\tau) | \psi_T \rangle$ which allows us to use the importance sampled distribution $f = \psi_\text{DMC}(\tau)\psi_T$ for our Monte Carlo estimates.
Details for calculating reduced density matrix elements in FN-DMC can be found in Wagner.

\pagebreak
\subsection{Defining $\mathcal{LE}$ for CuO molecule}
Based on anion photoelectron spectroscopy (APES) experiments\textbf{[ref]} on the CuO molecule we define our low-energy space $\mathcal{LE}$ as the space of states with 9 to 10 electrons in the Cu 3d orbitals. 
Mathematically we can express this space as the span of some eigenstates within our Hilbert space
\begin{equation}
\mathcal{LE} = \text{Span(}\{ |\Psi \rangle | \langle \Psi | n_{3d} | \Psi \rangle \ge 9,\ H|\Psi\rangle = E |\Psi\rangle \}\text{)}.
\end{equation}
The APES measurements indicate that the lowest energy excitations in the CuO molecule are between the O p-orbitals and Cu 4s orbitals, leaving around 10 electrons in the Cu 3d orbitals. 
Bonding between the O p and Cu 3d orbitals lead to significant variations in the Cu 3d occupation, indicating the importance of including excitations out of the Cu 3d orbitals. 
This is corroborated by APES measurements indicating that the lowest energy eigenstates with 9 electrons in the Cu 3d orbitals are only $\sim$ 0.2 eV higher in energy than the second excited state of the system with 10 electrons in the Cu 3d orbitals. 
Our particular focus will be on describing most accurately the properties and energies of eigenstates up to 2eV, as higher energy states form a nearly continuous block of states which include vibrational modes.

@@
